---
title: "Joining Data Sets with Merge()"
output: html_document
date: "2022-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Joining Data Sets with the Merge() Function Course Subsection

### Data Frames

Data frames provide more flexibility than matrices by allowing the mixing of data types.

#### Create a data frame:
```{r}
customers <- data.frame(id = c(1, 2, 3), name = c('Ann', 'Bill', 'Charles'), is.vip = c(TRUE, FALSE, TRUE))
customers

is.data.frame(customers)

head(customers)

str(customers)

# Note that name is a 'Factor w/ 3 levels"...
```
This means it is a variable that can take a limited number of variables/categories. This categorization makes sense in some situations but not here thus we must correct this:
```{r}
customers <- data.frame(id = c(1, 2, 3), name = c('Ann', 'Bill', 'Charles'), is.vip = c(TRUE, FALSE, TRUE), stringsAsFactors = FALSE)
customers

str(customers)
# Note that the name attribute is now a character vector
```

#### Loading data from a CSV file into a dataframe:
```{r}
customers <- read.csv('./data/M3. Joining Data Sets with the merge() function/C1. Data Frames/customers.csv')
customers

str(customers)
# Note we again run into the issue with attributes being incorrectly assigned the class of Factor

# The Fix:
customers <- read.csv('./data/M3. Joining Data Sets with the merge() function/C1. Data Frames/customers.csv', as.is = TRUE)
customers

str(customers)
```
To Recap:

* Data Frames can mix attribute types but...
* All data in a column must be of the same type
* Keep in mind, by default string variables can be assigned the type Factor which may be inaccurate

### Inner Joins

#### Example

* Data arrives in the form of two data frames:
    + First data frame contains customer names
    + Second data frame contains customer countries
```{r}
customer.names <- data.frame(id = 1:3, name = c('Helena', 'Astrid', 'Eduardo'))
customer.countries <- data.frame(customer_id = 2:4, country = c('USA', 'Spain', 'USA'))
customer.names
customer.countries

# Note there is some overlap between ids in the two data frames
```
The Problem:

* How can we merge these two data frames to get only rows with common ids?

An incorrect attempt: cbind()
```{r}
cbind(customer.names, customer.countries)
```
With cbind(), columns from both data frames are placed side by side and ids are mismatched.

* cbind() can only work if we have the same ids in the same order in each of the data frames.

#### Inner join with merge()

The solution:

* An inner join between the two data frames by the common variable: customer id

![An Inner Join Illustration Image](./data/M3. Joining Data Sets with the merge() function/C2. Inner Joins/inner_join.png)

```{r}
# Note: ?merge for documentation

merge(customer.names, customer.countries)
```
Note that merging them directly returns all possible combinations of rows between the two data frames:

* Three rows in each data frame - thus 3x3 = 9 rows, however...
* Most rows are invalid combinations
* To do a valid inner join, we need to explicitly state which common columns are meant to be involved

```{r}
merge(customer.names, customer.countries, by.x = 'id', by.y = 'customer_id')
```
Thus, merging these two columns gives us the inner-join of the two data frames.

#### Good Habits

* Use unique ids
    + But what if rows are uniquely identified by two columns instead of one... a composite unique id??
    + Good news, you can use a composite unique id:
```{r}
names <- data.frame(first.name = c('Helena', 'Helena', 'Tom'), last.name = c('Jones', 'Troy', 'Jones'), age = c(20, 30, 40))
countries <- data.frame(first.name = c('Helena', 'Helena', 'Tom'), last.name = c('Jones', 'Troy', 'Jones'), country = c('USA', 'UK', 'USA'))

# Incorrect Merge
merge(names, countries, by.x = 'first.name', by.y = 'first.name')

# Correct Merge
merge(names, countries, by.x = c('first.name', 'last.name'), by.y = c('first.name', 'last.name'))
```

A Final point on good habits for inner joins:

* Natural Joins can be useful, but their use should be limited to avoid errors:
    + A Natural Join entails relying on column names for joining...
    + It is the default used by the merge() function call
    + An example of a natural join based on the two data frames previously used:
    
```{r}
merge(names, countries)
```

What's happening here / why don't we need to be explicit??

* By default, merge uses the intersection of column names (reference documentation: ?merge).
    + Since both data frames have a first.name and last.name column, we do not need to be explicit.
* This works out in this scenario, but with larger data frames it can become messy and should be used sparingly.

A middle-ground:

* Merge and use the by parameter, specifying the columns:
```{r}
merge(names, countries, by = c('first.name', 'last.name'))
```
Note, this works because we are explicitly choosing columns present in both data frames.

Recap:

* Inner joins are a great way to pull data with a common unique identifier
* What about grabbing the identifiers that are not common? Can they also be retrieved?
    + This is answered in the Left, Right, and Full Outer Joins Section.








### Left, Right, and Full Outer Joins

### Data Frame Keys

### Data Frame Relationships










































































