---
title: "Managing Vectors"
output: html_document
date: "2022-09-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Managing Vectors Course Subsection

### Creating a Vector from Values

```{r cvfv}
# Code pairs with Creating a Vector from Values subsection of PS course

# Produce a vector with consecutive numbers
x <- 1:5
y <- 6:10
is.vector(x)
typeof(y)

plot(x, y)


```

A Vector is a conglomeration of values of the same type. 
Order Matters.
A number/value is a vector of 1.

```{r}
countries <- c('USA', 'UK')
typeof(countries)
length(countries)

# to check length of a string, use nchar() func
```

Combine function -c() is used when vector is unordered. 

All elements in a vector must be of the same type. If you want to combine elements of different types using -c(), R will use implicit coercion...
  changing types automatically to convert them to the same type

```{r}
mix <- c(1, TRUE, 'way')
mix
typeof(mix)
```

Nesting vectors inside of other vectors will flatten them:
```{r}
nesting_doll <- c(1, 2, c(4, 5, c(6, 'seven')))
nesting_doll
# Note that implicit coercion still applies
```


### Merging Values into a Vector

#### Merging Values at the start/end of a vector:
```{r}
v <- c(2:9)

# Merge values at the start
v <- c(1, v)

# Merge values at the end
v <- c(v, 10)

# Merging multiple values
v1 <- c(1:5)
v2 <- c(6:10)

v3 <- c(v1, v2)
v4 <- c(v2, v1)

v5 <- c(v1, 11:15)
```


#### Merging Values inside a vector -> to do so use append() func
```{r}
poem <- c('Mary', 'little', 'lamb')

poem <- append(poem, c('had', 'a'), after = 1)
```

Can see a behind the scenes look at append() func implementation:
```{r}
append
```
Here we see that the append() is a wrapper around the c() func.

#### Performance Implications of one big merge vs many small merges

One big merge is much better unless using a small sample size as the vector needs to be resized with every append.

### Merging Vectors into a Character Vector

Applies to scenarios such as merging strings or an sql query
```{r}
# Wrong way:
"hello" + "world"
```
Right way:
Use paste() function to:
  - Merge vectors of length = 1
  - Merge vectors of same length > 1
  - Merge vectors into one string
  
```{r}
# Recap
length("hello") # Vector of length 1
nchar("hello") # Composed of 5 characters
c('h', 'e', 'l', 'l', 'o') # Vector of length 5

# Merge vectors of length = 1
paste('hello', 'world') # Defaults to incorporating a space between elements

# Note: One or more objects can be pasted, to then be converted to a character vector (implicit coercion)
paste(1, 'two', TRUE)

# Test: Replace default separator with an empty space
paste('hello', 'world', sep = '')

# Merge vectors of same length > 1
paste(c('name', 'age'), c('John', 5), c('Doe', 'years')) # Returns a character vector of length 2: "name John Doe" "age 5 years"

# To merge vectors into one string... use collapse() arg
paste(c('name', 'age'), 
      c('John', 5), 
      c('Doe', 'years'), collapse = '-') # Returns: "name John Doe-age 5 years"
```

Also use paste() function and recycling to:
  - Merge vectors of different lengths
  
  - Recycling will automatically repeat or 'recycle' the shorter vectors to match the length of the largest vector
```{r}
paste(c('name', 'age', 'Name', 'Age'),
      c('John', 5)) # Returns "name John" "age 5"     "Name John" "Age 5" *Equivalent to copying and pasting the smaller vector seen below

paste(c('name', 'age', 'Name', 'Age'),
      c('John', 5,     'John', 5))

paste(c('name', 'age', 'Name', 'Age', 'Other'),
      c('John', 5,     'John', 5))

# Recycling is applicable to functions other than paste()... for ex:
c(1, 2) + 3 # Returns: [1] 4 5

# Which is equivalent to:
c(1, 2) +
c(3, 3) # Returns [1] 4 5

```

































































