---
title: "Managing Vectors"
output:
  html_document:
  pdf_document: 
date: "2022-09-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Managing Vectors Course Subsection

### Creating a Vector from Values

```{r cvfv}
# Produce a vector with consecutive numbers
x <- 1:5
y <- 6:10
x
y
is.vector(x)
typeof(y)

plot(x, y)


```

* A Vector is a conglomeration of values of the same type. 
* Order Matters.
* A number/value is a vector of 1.

```{r}
countries <- c('USA', 'UK')
typeof(countries) # Result: character
length(countries) # Result: 2

# to check length of a string, use nchar() func
```

Combine function -c() is used when vector is un-ordered. 

All elements in a vector must be of the same type. If you want to combine elements of different types using -c(), R will use implicit coercion... changing types automatically to convert them to the same type.

```{r}
mix <- c(1, TRUE, 'way')
mix
typeof(mix)
```

Nesting vectors inside of other vectors will flatten them:
```{r}
nesting_doll <- c(1, 2, c(4, 5, c(6, 'seven')))
nesting_doll
# Note that implicit coercion still applies
```


### Merging Values into a Vector

#### Merging Values at the start/end of a vector:
```{r}
v <- c(2:9)

# Merge values at the start
v <- c(1, v)
v

# Merge values at the end
v <- c(v, 10)
v

# Merging multiple values
v1 <- c(1:5)
v2 <- c(6:10)

v3 <- c(v1, v2)
v4 <- c(v2, v1)

v5 <- c(v1, 11:15)

v3
v4
v5
```


#### Merging Values inside a vector is accomplished through the use of the append() func
```{r}
poem <- c('Mary', 'little', 'lamb')

poem <- append(poem, c('had', 'a'), after = 1)
poem
```

Can see a behind the scenes look at append() func implementation:
```{r}
append
```
Here we see that the append() is a wrapper around the c() func.

#### Performance Implications of one big merge vs many small merges

One big merge is much better unless using a small sample size as the vector needs to be resized with every append.

### Merging Vectors into a Character Vector

Applies to scenarios such as merging strings or an sql query
```{r}
# Wrong way:
#"hello" + "world"
```
Right way:
Use paste() function to:

* Merge vectors of length = 1
* Merge vectors of same length > 1
* Merge vectors into one string
  
```{r}
# Recap
length("hello") # Vector of length 1
nchar("hello") # Composed of 5 characters
c('h', 'e', 'l', 'l', 'o') # Vector of length 5

# Merge vectors of length = 1
paste('hello', 'world') # Defaults to incorporating one space between elements

# Note: One or more objects can be pasted, to then be converted to a character vector (implicit coercion)
paste(1, 'two', TRUE)

# Test: Replace default separator with an empty space
paste('hello', 'world', sep = '')

# Merge vectors of same length > 1
paste(c('name', 'age'), c('John', 5), c('Doe', 'years')) # Returns a character vector of length 2: "name John Doe" "age 5 years"

# To merge vectors into one string... use collapse() arg
paste(c('name', 'age'), 
      c('John', 5), 
      c('Doe', 'years'), collapse = '-') # Returns: "name John Doe-age 5 years"
```

Also use paste() function and recycling to:

* Merge vectors of different lengths
* Recycling will automatically repeat or 'recycle' the shorter vectors to match the length of the largest vector

```{r}
paste(c('name', 'age', 'Name', 'Age'),
      c('John', 5)) # Returns "name John" "age 5"     "Name John" "Age 5" *Equivalent to copying and pasting the smaller vector seen below

paste(c('name', 'age', 'Name', 'Age'),
      c('John', 5,     'John', 5))

paste(c('name', 'age', 'Name', 'Age', 'Other'),
      c('John', 5,     'John', 5))

# Recycling is applicable to functions other than paste()... for ex:
c(1, 2) + 3 # Returns: [1] 4 5

# Which is equivalent to:
c(1, 2) +
c(3, 3) # Returns [1] 4 5

```

#### Merging Vectors into a Matrix

What is a matrix in R? 

* A vector with multiple dimensions instead of 1.
* All values in a matrix must be of the same type.

```{r}
first_matrix <- matrix(1, nrow = 2, ncol = 3)
first_matrix

is.matrix(first_matrix) # Returns: TRUE
is.vector(first_matrix) # Returns: FALSE
typeof(first_matrix) # Returns: double

length(first_matrix) # Returns: 6

larger_matrix <- matrix(1:9, nrow = 3, ncol = 3)
larger_matrix

# Isolate values from a matrix using row and col indicies
second_col <- larger_matrix[,2]
second_col

is.vector(second_col) # Returns TRUE

third_row <- larger_matrix[3,]
third_row

is.vector(third_row) # Returns TRUE

```

Thus, a matrix is composed of a series of vectors stacked on top of one another, or a series of columns placed side by side.
We should logically then be able to build out a matrix by stacking rows or adding columns.
To do so, we use cbind()/rbind() functions to merge vectors into a matrix:

```{r}
# Combining vectors of length 1
cbind('hello', 'world')

rbind('hello', 'world')

# Combining vectors of length > 1
cbind(c(1, 2, 3), c(4, 5, 6)) # Returns: a 3x2 matrix (cbind "sees" two distinct objects)
rbind(c(1, 2, 3), c(4, 5, 6)) # Returns: a 2x3 matrix

#Combining vectors of different lengths - Recycling rule is again applied
cbind(c(1, 2, 3), 5)
rbind(c('a', 'b', 'c', 'd', 'e'), c(1, 2), TRUE)
# Note that implicit coercion is applied
```


